---
title: "MSstats — Differential Abundance Analysis"
---

## Libraries

```{r setup, include=TRUE, message=FALSE}
library(data.table)
library(MSstats)
library(qs2)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(tidyverse)
library(readr)
```

## Set directories

```{r}
base_dir <- "/nemo/lab/destrooperb/home/shared/zanettc/giulia_proteomics"
run_num  <- "run3"

results_dir <- file.path(base_dir, "results", run_num)
objects_dir <- file.path(base_dir, "data", "processed", run_num)
input_dir   <- file.path(base_dir, "input")

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(objects_dir, recursive = TRUE, showWarnings = FALSE)
```

## Build annotation from study design

All 16 samples passed QC. R.Condition = "Not Defined" in the raw file because the study
design was not configured inside Spectronaut — we define it here from the known design.

```{r}
msstats_annotation <- data.table(
  Run          = paste0("GA_", 1:16),
  Condition    = rep(c("hApp", "hAppNLGF", "hAppNLGF_FIRE", "hApp_FIRE"), each = 4),
  BioReplicate = rep(1:4, times = 4)
)
msstats_annotation

fwrite(msstats_annotation, file.path(objects_dir, "msstats_annotation.csv"))
```

## Load raw MSstats Spectronaut proteome report

Only the columns required by MSstats are read — this keeps memory manageable on the 8.4 GB file.

```{r}
msstats_cols <- c(
  "R.Condition", "R.FileName", "R.Replicate",
  "PG.ProteinGroups", "PG.ProteinAccessions",
  "PEP.StrippedSequence",
  "EG.ModifiedSequence", "EG.PrecursorId", "EG.Qvalue",
  "FG.Charge", "FG.Id",
  "F.FrgIon", "F.Charge", "F.FrgLossType", "F.FrgNum", "F.FrgType",
  "F.NormalizedPeakArea", "F.PeakArea", "F.ExcludedFromQuantification"
)

cat("Reading Spectronaut MSstats report...\n")
raw_spectronaut <- fread(
  file.path(input_dir, "20260423_154110_PCF000612_GA_MSStats_Report-Proteome.csv"),
  select = msstats_cols
)
cat("Rows loaded:", nrow(raw_spectronaut), "\n")

qs_save(raw_spectronaut, file.path(objects_dir, "msstats_report.qs2"))
```

## Reload from .qs2 (skip the read step on re-runs)

```{r}
#raw_spectronaut <- qs_read(file.path(objects_dir, "msstats_report.qs2"))
```

## Overwrite R.Condition and R.Replicate with study design values

SpectronauttoMSstatsFormat uses R.Condition internally before merging with the annotation.
We replace "Not Defined" with the correct values from the annotation to avoid any issues.

```{r}
raw_spectronaut[msstats_annotation,
                on = c("R.FileName" = "Run"),
                `:=`(R.Condition = i.Condition,
                     R.Replicate = i.BioReplicate)]

cat("Condition distribution after fix:\n")
print(raw_spectronaut[, .N, by = .(R.Condition, R.FileName)][order(R.Condition, R.FileName)])
```

## Amyloid-β tryptic peptide verification

Confirms that hApp and hAppNLGF conditions have detectable Aβ signal.
In microglia, signal reflects phagocytosed / secreted Aβ rather than endogenous APP expression.

```{r}
arctic_peptide <- "LVFFAGDVGSNK"
normal_peptide <- "LVFFAEDVGSNK"
peptide_both <- "DAEFRHDSGYEVHHQK"

target_peptide <- "LVFFAEDVGSNK"  # Conserved Aβ tryptic peptide (mouse and human APP)



ab_check <- raw_spectronaut %>%
  filter(PEP.StrippedSequence == target_peptide) %>%
  group_by(R.FileName, R.Condition) %>%
  summarise(Total_Intensity = sum(F.PeakArea, na.rm = TRUE), .groups = "drop") %>%
  mutate(Log2_Intensity = log2(Total_Intensity + 1)) %>%
  # Extract numeric sample ID for proper sorting
  mutate(Sample_Num = as.numeric(gsub("GA_", "", R.FileName))) %>%
  arrange(R.Condition, Sample_Num) %>%
  mutate(R.FileName = factor(R.FileName, levels = unique(R.FileName)))

if (nrow(ab_check) == 0) {
  cat("Aβ peptide (", target_peptide, ") not detected in any sample.\n")
} else {
  cat("Aβ peptide detected in", nrow(ab_check), "sample(s):\n")
  print(ab_check)
}

p_ab <- ggplot(ab_check, aes(x = R.FileName, y = Log2_Intensity, fill = R.Condition)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c(
    "hApp" = "#1b9e77", "hAppNLGF" = "#d95f02",
    "hAppNLGF_FIRE" = "#7570b3", "hApp_FIRE" = "#e7298a"
  )) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Aβ Tryptic Peptide Signal per Sample",
       subtitle = paste0("Peptide: ", target_peptide),
       x = "Sample", y = "Log2 Intensity")

ggsave(file.path(results_dir, "AB_Peptide_Verification.pdf"), p_ab, width = 10, height = 6)
print(p_ab)
```

## Convert to MSstats input format

Feature-level filtering: Qvalue < 0.01, remove features with too few measurements.

```{r}
msstats_input <- suppressWarnings(SpectronauttoMSstatsFormat(
  raw_spectronaut,
  annotation                  = msstats_annotation,
  filter_with_Qvalue          = TRUE,
  qvalue_cutoff               = 0.01,
  fewMeasurements             = "remove",
  removeProtein_with1Feature  = FALSE
))

cat("MSstats input rows:", nrow(msstats_input), "\n")
cat("Unique proteins:   ", uniqueN(msstats_input$ProteinName), "\n")

qs_save(msstats_input, file.path(objects_dir, "msstats_input.qs2"))
```

## Feature completeness filter

Keeps features observed in ≥50% of runs within at least one condition.
This removes near-absent features that destabilise the linear model. 
This is the same as saying proteins must be present in at least 2 samples out 
of 3 for at least one condition. 

```{r}
#Read in if starting here
msstats_input <- qs_read(file.path(base_dir, "data", "processed", "run1", "msstats_input.qs2"))

msstats_input_filt <- as.data.frame(msstats_input) %>%
  group_by(ProteinName, PeptideSequence, PrecursorCharge, FragmentIon, ProductCharge, Condition) %>%
  summarise(
    n_obs   = sum(!is.na(Intensity) & Intensity > 0),
    n_total = n(),
    prop    = n_obs / n_total,
    .groups = "drop"
  ) %>%
  group_by(ProteinName, PeptideSequence, PrecursorCharge, FragmentIon, ProductCharge) %>%
  filter(any(prop >= 0.5)) %>%
  ungroup() %>%
  inner_join(
    as.data.frame(msstats_input),
    by = c("ProteinName", "PeptideSequence", "PrecursorCharge",
           "FragmentIon", "ProductCharge", "Condition")
  )

message("Rows before completeness filter: ", nrow(msstats_input))
message("Rows after  completeness filter: ", nrow(msstats_input_filt))
message("Proteins retained: ", uniqueN(msstats_input_filt$ProteinName))

table(
  Value_Type = case_when(
    is.na(msstats_input$Intensity) ~ "True NAs",
    msstats_input$Intensity == 0 ~ "Exact Zeros (Filtered by Q-value)",
    msstats_input$Intensity > 0 ~ "Valid Measurements"
  )
)
```

## (Optional) Keratin protein filter

The MSstats report contains only UniProt accessions in ProteinName — no gene symbols.
To filter keratins, build a lookup from UniProt or biomaRt and enable the block below.

```{r}
## Keratin protein filter

protein_dict <- read_csv(file.path(base_dir, "data", "processed", "run1", "protein_dictionary.csv"))

# Pull all UniProt accessions annotated as keratin (Krt* gene symbol).
# The Protein column can be a semicolon-separated group, so split before matching.
keratin_ids <- protein_dict %>%
  filter(str_detect(Gene, "^Krt") & !str_detect(Gene, "^Krtcap")) %>%  # drop Krtcap2 (not a keratin contaminant)
  pull(Protein) %>%
  str_split(";") %>%
  unlist() %>%
  unique()

message("Keratin accessions to filter: ", length(keratin_ids))

# Match against ProteinName by splitting protein groups the same way,
# so a group is dropped if ANY member is a keratin.
msstats_input_filt <- msstats_input_filt %>%
  filter(!map_lgl(str_split(ProteinName, ";"),
                  ~ any(.x %in% keratin_ids)))

message("Rows after keratin filter: ", nrow(msstats_input_filt))
message("Proteins after keratin filter: ", n_distinct(msstats_input_filt$ProteinName))

```

## Data summarisation and normalisation

Summarises fragment intensities to protein level with TMP, median normalisation.

```{r}
processed_data <- suppressWarnings(dataProcess(
  msstats_input_filt,
  normalization                        = "equalizeMedians",
  summaryMethod                        = "TMP",
  censoredInt                          = 'NA',
  MBimpute                             = TRUE,
  featureSubset                        = "highQuality",
  remove_uninformative_feature_outlier = TRUE
))

qs_save(processed_data, file.path(objects_dir, "processed_msstats_data.qs2"))
```

## Reload processed data (if re-running from this point)

```{r}
#processed_data <- qs_read(file.path(objects_dir, "processed_msstats_data.qs2"))
```

## Verify condition levels

MSstats sorts conditions alphabetically — confirm the order before building the contrast matrix.

```{r}
condition_levels <- levels(processed_data$ProteinLevelData$GROUP)
cat("Condition levels (alphabetical order used in contrast matrix):\n")
print(condition_levels)
# Expected: "hApp"  "hApp_FIRE"  "hAppNLGF"  "hAppNLG_FIRE"
# (underscore sorts after capital letters in most R locales,
#  but verify with the output above before proceeding)
```

## Build contrast matrix

Comparisons (numerator vs denominator):

| Contrast | Question |
|---|---|
| hAppNLGF vs hApp | NLGF mutation effect in microglia-containing condition |
| hApp_FIRE vs hApp | Microglia depletion effect on hApp background |
| hAppNLG_FIRE vs hAppNLGF | Microglia depletion effect on NLGF background |
| hAppNLG_FIRE vs hApp_FIRE | NLGF mutation effect in microglia-depleted condition |

```{r}
# Helper: builds a named contrast vector given the full condition_levels vector
make_contrast <- function(numerator, denominator, levels) {
  v <- setNames(rep(0L, length(levels)), levels)
  v[numerator]   <-  1L
  v[denominator] <- -1L
  v
}

comparison_matrix <- rbind(
  make_contrast("hAppNLGF",     "hApp",         condition_levels),
  make_contrast("hApp_FIRE",    "hApp",         condition_levels),
  make_contrast("hAppNLGF_FIRE", "hAppNLGF",     condition_levels),
  make_contrast("hAppNLGF_FIRE", "hApp_FIRE",    condition_levels)
)

rownames(comparison_matrix) <- c(
  "hAppNLGF_vs_hApp",
  "hApp_FIRE_vs_hApp",
  "hAppNLGF_FIRE_vs_hAppNLGF",
  "hAppNLGF_FIRE_vs_hApp_FIRE"
)

print(comparison_matrix)
```

## Statistical modelling

Linear mixed-effects model across all proteins.

```{r}
test_results <- suppressWarnings(groupComparison(contrast.matrix = comparison_matrix, data = processed_data))

qs_save(test_results, file.path(objects_dir, "comparison_results.qs2"))

res_df <- as.data.table(test_results$ComparisonResult)
cat("Total protein-comparison rows:", nrow(res_df), "\n")
cat("Significant (adj.pvalue < 0.05):",
    nrow(res_df[adj.pvalue < 0.05, ]), "\n")

fwrite(res_df, file.path(results_dir, "Full_GroupComparison_Results.csv"))
cat("Results saved to:", file.path(results_dir, "Full_GroupComparison_Results.csv"), "\n")
```

## Quick summary of results

```{r}
res_df %>%
  filter(!is.na(adj.pvalue)) %>%
  group_by(Label) %>%
  summarise(
    n_tested = n(),
    n_sig    = sum(adj.pvalue < 0.05, na.rm = TRUE),
    n_up     = sum(adj.pvalue < 0.05 & log2FC > 0, na.rm = TRUE),
    n_down   = sum(adj.pvalue < 0.05 & log2FC < 0, na.rm = TRUE),
    .groups  = "drop"
  ) %>%
  arrange(Label)
```

## Reload comparison results (if re-running from this point)

```{r}
#test_results <- qs_read(file.path(objects_dir, "comparison_results.qs2"))
#res_df <- as.data.table(test_results$ComparisonResult)
```
