---
title: "1. Phosphoproteome — Initial QC"
---

## Overview

QC for the phosphoproteome MSstats report from Spectronaut. Mirrors the bulk-proteome QC (notebook `01_initial_qc_spectronaut.qmd` in `bulk_proteomics/`) but operates on phosphopeptide-level data and tracks phosphosite-level metrics.

**Important note on this Spectronaut export:** the file does not include `EG.PTMLocalizationProbabilities` (or `EG.PTMAssayProbability`). All phospho-site information lives in `EG.ModifiedSequence` (e.g. `_AAS[Phospho (STY)]TPGK_`). MSstatsPTM downstream needs a mouse FASTA to map sequence positions to protein residues — see notebook 02. For QC here we count phosphosites by parsing `EG.ModifiedSequence` directly.

## Libraries

```{r setup, include=FALSE, message=TRUE}
library(data.table)
library(qs2)
library(dplyr)
library(ggplot2)
library(stringr)
library(tidyr)
library(pheatmap)
```

## Directories

```{r}
base_dir <- "/nemo/lab/destrooperb/home/shared/zanettc/giulia_proteomics/phosphoproteomics"
run_num  <- "run2"

results_dir <- file.path(base_dir, "results", run_num)
objects_dir <- file.path(base_dir, "data", "processed", run_num)
input_dir   <- file.path(base_dir, "input")

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(objects_dir, recursive = TRUE, showWarnings = FALSE)
```

## Study design

Same 16-sample 2×2 design as the bulk proteome.

```{r}
meta <- data.table(
  FileName      = paste0("GA_", 1:16),
  Condition     = rep(c("hApp", "hAppNLGF", "hAppNLGF_FIRE", "hApp_FIRE"), each = 4),
  BioReplicate  = rep(1:4, times = 4),
  Has_Microglia = rep(c(TRUE, TRUE, FALSE, FALSE), each = 4)
)
meta
```

## Load the phospho report

Read only the columns needed for QC and phosphosite parsing.

```{r load}
phospho_cols <- c(
  "R.Condition", "R.FileName", "R.Replicate",
  "PG.ProteinGroups", "PG.ProteinAccessions", "PG.Qvalue", "PG.Quantity",
  "PEP.StrippedSequence",
  "EG.ModifiedSequence", "EG.PrecursorId", "EG.Qvalue",
  "FG.Charge", "FG.Id",
  "F.FrgIon", "F.Charge", "F.FrgLossType", "F.FrgNum", "F.FrgType",
  "F.NormalizedPeakArea", "F.PeakArea", "F.ExcludedFromQuantification"
)

cat("Reading phospho MSstats report...\n")
phospho_raw <- fread(
  file.path(input_dir, "PCF000612_phospho_GA_MSStats_report.csv"),
  select = phospho_cols
)
cat("Rows:", nrow(phospho_raw), "\n")
cat("Unique samples:", uniqueN(phospho_raw$R.FileName), "\n")
cat("Samples:", paste(sort(unique(phospho_raw$R.FileName)), collapse = ", "), "\n")
```

## Restrict to confidently-identified phospho-modified precursors

Filter to phospho-containing precursors at EG.Qvalue < 0.01.

```{r filter}
is_phospho <- str_detect(phospho_raw$EG.ModifiedSequence, fixed("Phospho"))
cat("Phospho-modified precursor rows: ",
    sum(is_phospho), "/", nrow(phospho_raw), "\n")

phospho_raw <- phospho_raw[is_phospho & EG.Qvalue < 0.01]
cat("After phospho + Qvalue<0.01 filter:", nrow(phospho_raw), "\n")
```

## Phosphopeptide and phosphosite metrics per sample

A "phosphosite" here is a (protein, peptide, S/T/Y position-within-peptide) tuple — the *peptide-relative* position. Absolute protein-residue numbering requires the FASTA mapping, done in notebook 02.

```{r site_metrics}
# Helper: extract phospho positions within the modified sequence (Vectorized)
extract_phospho_positions_vec <- function(modseq) {
  
  # 1. Remove leading/trailing underscores
  m <- str_replace_all(modseq, "_", "")
  
  # 2. Strip non-phospho mods entirely. 
  # Use ifelse() so it handles vectors of length > 1 correctly
  m_clean <- str_replace_all(m, "\\[[^\\]]*\\]", function(x) {
    ifelse(str_detect(x, "Phospho"), x, "")
  })
  
  # 3. Split the sequence on the exact Phospho tags
  parts <- str_split(m_clean, "\\[Phospho \\(STY\\)\\]")
  
  # 4. Count the residues and calculate positions
  vapply(parts, function(p) {
    # If the length is 1, it means no split occurred (no phospho present or NA)
    if (length(p) == 1) return(NA_character_)  
    
    # Each part before the last contains only residue chars
    residue_counts <- nchar(p[-length(p)])
    positions <- cumsum(residue_counts)
    
    # Return as a semicolon-separated string (e.g., "3;5")
    paste(positions, collapse = ";")
  }, FUN.VALUE = character(1)) 
}

cat("Parsing phosphosite positions (vectorized)...\n")

# Vectorized call — completely replaces mapply + unlist
phospho_raw[, PhosphoPos := extract_phospho_positions_vec(EG.ModifiedSequence)]

# Per-sample summary -> need to use correct filenames
phospho_raw[, R.FileName := str_replace(R.FileName, "^GA_P_", "GA_")]

# Then re-run sample_summary as before
sample_summary <- phospho_raw[
  , .(
    n_features        = uniqueN(EG.PrecursorId),
    n_modseq          = uniqueN(EG.ModifiedSequence),
    n_phosphopeptides = uniqueN(paste(PG.ProteinGroups, EG.ModifiedSequence)),
    n_phosphoproteins = uniqueN(PG.ProteinGroups)
  ),
  by = R.FileName
]
sample_summary <- sample_summary[meta, on = c(R.FileName = "FileName")]
sample_summary <- sample_summary[meta, on = c(R.FileName = "FileName")]

print(sample_summary[order(R.FileName)])
fwrite(sample_summary, file.path(results_dir, "phospho_sample_summary.csv"))

unique(phospho_raw$R.FileName)

```

## Phosphopeptide count per sample (bar plot)

```{r plot_counts, fig.width=10, fig.height=5}
sample_summary[, R.FileName := factor(R.FileName, levels = paste0("GA_", 1:16))]

p_counts <- ggplot(sample_summary,
                   aes(x = R.FileName, y = n_phosphopeptides, fill = Condition)) +
  geom_col() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Phosphopeptides identified per sample",
       x = NULL, y = "# phosphopeptides (proteinGroup × modSeq)") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_counts)
ggsave(file.path(results_dir, "Phosphopeptides_per_sample.pdf"),
       p_counts, width = 10, height = 5)
```

## MAD-based outlier flag on phosphopeptide count

```{r mad_outliers}
mad_thresh <- 3
med <- median(sample_summary$n_phosphopeptides)
md  <- mad(sample_summary$n_phosphopeptides)

outliers <- sample_summary[abs(n_phosphopeptides - med) > mad_thresh * md]
cat("Median:", med, " MAD:", md, "\n")
cat("Outlier samples (>", mad_thresh, "× MAD):\n")
print(outliers[, .(R.FileName, Condition, n_phosphopeptides)])
```

## PCA on phosphopeptide intensities

Aggregate to one log2 intensity per (modSeq × sample), wide format, log2-transform, replace NAs with column-wise medians for PCA only (this is QC, not stats).

```{r pca, fig.width=8, fig.height=6}
# One value per (proteinGroup, modSeq, sample) — sum fragment areas
phospho_intens <- phospho_raw[
  , .(Intensity = sum(F.PeakArea, na.rm = TRUE)),
  by = .(PG.ProteinGroups, EG.ModifiedSequence, R.FileName)
]

# Drop near-zero intensities
phospho_intens <- phospho_intens[Intensity > 0]
phospho_intens[, log2I := log2(Intensity + 1)]

wide <- dcast(phospho_intens,
              PG.ProteinGroups + EG.ModifiedSequence ~ R.FileName,
              value.var = "log2I")
mat <- as.matrix(wide[, -(1:2)])
rownames(mat) <- paste(wide$PG.ProteinGroups, wide$EG.ModifiedSequence)

# Keep features observed in >= 12 of 16 samples for PCA stability
keep <- rowSums(!is.na(mat)) >= 12
mat <- mat[keep, , drop = FALSE]

# Median-impute NAs by column for PCA only
for (j in seq_len(ncol(mat))) {
  miss <- is.na(mat[, j])
  if (any(miss)) mat[miss, j] <- median(mat[, j], na.rm = TRUE)
}

pca <- prcomp(t(mat), scale. = TRUE)
pca_df <- as.data.frame(pca$x[, 1:3])
pca_df$Sample    <- rownames(pca_df)
pca_df$Condition <- meta$Condition[match(pca_df$Sample, meta$FileName)]

ve <- summary(pca)$importance[2, ] * 100

p_pca <- ggplot(pca_df, aes(PC1, PC2, color = Condition, label = Sample)) +
  geom_point(size = 4) +
  ggrepel::geom_text_repel(size = 3, max.overlaps = 50) +
  scale_color_brewer(palette = "Set2") +
  labs(title = "PCA — phosphopeptide log2 intensity",
       subtitle = paste0("n features = ", nrow(mat)),
       x = sprintf("PC1 (%.1f%%)", ve[1]),
       y = sprintf("PC2 (%.1f%%)", ve[2])) +
  theme_minimal(base_size = 12)

print(p_pca)
ggsave(file.path(results_dir, "Phospho_PCA.pdf"), p_pca, width = 8, height = 6)
```

## Sample-correlation heatmap

```{r corr_heatmap, fig.width=8, fig.height=7}
cor_mat <- cor(mat, use = "pairwise.complete.obs", method = "spearman")

ann_col <- data.frame(Condition = meta$Condition,
                      row.names = meta$FileName)

ph <- pheatmap(cor_mat,
               annotation_col = ann_col,
               main = "Phospho — Spearman correlation between samples",
               fontsize = 9,
               filename = file.path(results_dir, "Phospho_correlation_heatmap.pdf"),
               width = 8, height = 7)
```

## Save cleaned phospho data for downstream notebooks

```{r save}
qs_save(phospho_raw, file.path(objects_dir, "phospho_raw_filtered.qs2"))
fwrite(sample_summary, file.path(objects_dir, "phospho_sample_summary.csv"))

cat("Saved:\n",
    "  ", file.path(objects_dir, "phospho_raw_filtered.qs2"), "\n",
    "  ", file.path(objects_dir, "phospho_sample_summary.csv"), "\n")
```
