---
title: "3. Phospho — DE volcanoes (adjusted)"
---

## Overview

Volcano plots and per-contrast site-level CSVs from the adjusted PTM contrasts in notebook 02. Adjusted log2FC = "phospho minus protein," i.e. regulation not explained by total protein abundance.

Sites are labelled in `Gene_Sxxx` format (e.g. `Mapt_S202`) when the FASTA mapping in notebook 02 succeeded; otherwise the raw protein/peptide identifier is used.

## Libraries

```{r setup, include=FALSE, message=FALSE}
library(data.table)
library(qs2)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(stringr)
library(AnnotationDbi)
library(org.Mm.eg.db)
```

## Directories

```{r}
base_dir <- "/nemo/lab/destrooperb/home/shared/zanettc/giulia_proteomics/phosphoproteomics"
run_num  <- "run2"

results_dir <- file.path(base_dir, "results", run_num)
objects_dir <- file.path(base_dir, "data", "processed", run_num)
dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
```

## Load adjusted PTM contrast table

```{r load}
res_df <- fread(file.path(results_dir, "PTM_adjusted_GroupComparison.csv"))
cat("Rows:", nrow(res_df), "  Contrasts:",
    paste(unique(res_df$Label), collapse = ", "), "\n")

# MSstatsPTM puts site identifier in `Protein` (e.g. UniProt_S202_T205); split out
# the UniProt accession to map to a gene symbol. The exact format depends on the
# FASTA-mapping success in notebook 02 — adjust the regex if your sites look
# different.
res_df[, Accession := str_extract(Protein, "^[^_;]+")]
res_df[, Sites_in_id := str_extract(Protein, "(?<=_)[STY][0-9]+(?:_[STY][0-9]+)*")]

# Map UniProt accession → gene symbol
res_df[, Gene := mapIds(
  org.Mm.eg.db,
  keys      = Accession,
  column    = "SYMBOL",
  keytype   = "UNIPROT",
  multiVals = "first"
)]
res_df[, Gene := ifelse(is.na(Gene) | Gene == "", Accession, Gene)]
res_df[, SiteLabel := ifelse(is.na(Sites_in_id), Gene, paste0(Gene, "_", Sites_in_id))]
```

## Volcano helper (mirrors bulk's `generate_swanky_volcano`)

```{r volcano_fn}
volcano_cols <- c("Upregulated"     = "firebrick3",
                  "Downregulated"   = "navy",
                  "Not Significant" = "grey80")

generate_phospho_volcano <- function(df, comparison_label,
                                     logfc_cutoff = 0.5,
                                     graphs_dir   = results_dir) {
  v <- df %>%
    filter(Label == comparison_label,
           !is.na(adj.pvalue), is.finite(log2FC)) %>%
    mutate(
      Status = case_when(
        adj.pvalue < 0.05 & log2FC >  logfc_cutoff ~ "Upregulated",
        adj.pvalue < 0.05 & log2FC < -logfc_cutoff ~ "Downregulated",
        TRUE ~ "Not Significant"
      )
    ) %>%
    group_by(Status) %>%
    mutate(GroupRank = rank(adj.pvalue, ties.method = "first")) %>%
    ungroup() %>%
    mutate(PlotLabel = ifelse(Status != "Not Significant" & GroupRank <= 20,
                               SiteLabel, NA_character_))

  p <- ggplot(v, aes(x = log2FC, y = -log10(adj.pvalue),
                     fill = Status, color = Status)) +
    geom_point(aes(size = Status, alpha = Status), shape = 21, stroke = 0.2) +
    scale_fill_manual(values  = volcano_cols) +
    scale_color_manual(values = c("Upregulated"     = "black",
                                   "Downregulated"   = "black",
                                   "Not Significant" = "grey90")) +
    scale_size_manual(values  = c("Upregulated" = 3, "Downregulated" = 3,
                                   "Not Significant" = 1.5)) +
    scale_alpha_manual(values = c("Upregulated" = 1, "Downregulated" = 1,
                                   "Not Significant" = 0.3)) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed",
               color = "grey30", linewidth = 0.8) +
    geom_vline(xintercept = c(-logfc_cutoff, logfc_cutoff), linetype = "dashed",
               color = "grey30", linewidth = 0.8) +
    geom_label_repel(aes(label = PlotLabel),
                     fill = "white", color = "black",
                     fontface = "bold", size = 3.2,
                     box.padding = 0.5, point.padding = 0.5,
                     min.segment.length = 0, segment.color = "grey50",
                     max.overlaps = 50) +
    labs(title    = paste0("Adjusted phospho DE — ", gsub("_", " ", comparison_label)),
         subtitle = paste0("Top 20 sites labelled | FDR < 0.05 | |log2FC| > ", logfc_cutoff,
                           " (phospho minus protein)"),
         x = bquote(bold("Adjusted log"[2] ~ "FC")),
         y = bquote(bold("-Log"[10] ~ "FDR"))) +
    theme_minimal(base_size = 14) +
    theme(plot.title    = element_text(face = "bold", size = 18, hjust = 0),
          plot.subtitle = element_text(size = 12, color = "grey40",
                                       margin = margin(b = 10)),
          axis.title    = element_text(face = "bold", size = 14),
          axis.text     = element_text(face = "bold", color = "black"),
          axis.line     = element_line(color = "black", linewidth = 1),
          axis.ticks    = element_line(color = "black", linewidth = 1),
          legend.position = "top",
          legend.title    = element_blank(),
          legend.text     = element_text(size = 12, face = "bold"),
          panel.grid.minor = element_blank(),
          panel.grid.major = element_line(color = "grey95"))

  safe <- gsub(" ", "_", comparison_label)
  ggsave(file.path(graphs_dir, paste0("PhosphoVolcano_", safe, ".pdf")),
         p, width = 10, height = 7, dpi = 300, device = cairo_pdf)

  fwrite(v %>% dplyr::select(SiteLabel, Gene, Accession, Sites_in_id,
                      log2FC, pvalue, adj.pvalue, Status) %>%
           arrange(adj.pvalue),
         file.path(graphs_dir, paste0("PhosphoVolcano_Data_", safe, ".csv")))
  p
}
```

## Iterate through contrasts

```{r run_volcanoes, fig.width=10, fig.height=7}
for (cl in unique(res_df$Label)) {
  print(generate_phospho_volcano(res_df, cl, logfc_cutoff = 0.5,
                                 graphs_dir = results_dir))
}
```

## Quick sanity check — bulk vs phospho concordance

For each contrast, scatter the protein-level log2FC against the unadjusted phospho log2FC at sites whose protein is present in both. Diagonal = abundance-driven. Off-diagonal = phosphorylation-specific regulation.

```{r concordance, fig.width=10, fig.height=8}
ptm_unadj <- fread(file.path(results_dir, "PTM_unadjusted_GroupComparison.csv"))
prot_res  <- fread(file.path(results_dir, "PROTEIN_GroupComparison.csv"))

# Map each phospho row's parent accession back to the protein result
ptm_unadj[, Accession := str_extract(Protein, "^[^_;]+")]
prot_res[, Accession := str_extract(Protein, "^[^;]+")]

cmp <- merge(
  ptm_unadj[, .(Label, SiteProtein = Protein, Accession,
                Phospho_log2FC = log2FC, Phospho_padj = adj.pvalue)],
  prot_res[, .(Label, Accession,
               Protein_log2FC = log2FC, Protein_padj = adj.pvalue)],
  by = c("Label", "Accession")
)

p_scat <- ggplot(cmp,
                 aes(Protein_log2FC, Phospho_log2FC,
                     color = Phospho_padj < 0.05)) +
  geom_point(alpha = 0.5, size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey30") +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey60") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey60") +
  scale_color_manual(values = c("TRUE" = "firebrick3", "FALSE" = "grey60"),
                     name   = "Phospho FDR < 0.05") +
  facet_wrap(~ Label, ncol = 3, scales = "free") +
  labs(title = "Phospho vs Protein log2FC per contrast",
       x = "Protein log2FC (bulk)",
       y = "Phospho log2FC (unadjusted)") +
  theme_bw(base_size = 11) +
  theme(legend.position = "top", strip.text = element_text(face = "bold"))

print(p_scat)
ggsave(file.path(results_dir, "Phospho_vs_Protein_log2FC_scatter.pdf"),
       p_scat, width = 12, height = 8, device = cairo_pdf)
```
```{r}
# 1. Diagnose: how many sites are at the sentinel values?
ptm_unadj[, .(.N), by = .(Label, log2FC)][order(-N)][1:20]
# If you see large N at specific extreme values, those are the sentinels.

# 2. Flag them explicitly rather than letting them dominate the plot
cmp[, imputed_phospho := abs(Phospho_log2FC) >= quantile(abs(Phospho_log2FC), 0.999, na.rm = TRUE) &
                         duplicated(Phospho_log2FC) | 
                         duplicated(Phospho_log2FC, fromLast = TRUE)]

# Better: MSstats sets a column `issue` for these. Check it:
unique(ptm_unadj$issue)
# Look for "oneConditionMissing" or similar — those are your sentinel rows.
```

