---
title: "4. Kinase activity inference"
---

## Overview

Inferred kinase activity per contrast from the adjusted phospho contrast results. Translates "site `X` of protein `Y` went up" into "kinase `K` is more active." Uses `decoupleR` with the OmniPath kinase–substrate network (mouse). Two methods are run for robustness:

- **WMEAN** — weighted mean of substrate t-statistics (fast, robust to network noise)
- **ULM** — univariate linear model per kinase (calibrated p-values)

Both produce per-kinase activity scores and FDR-adjusted significance per contrast.

## Setup notes
- The OmniPath query needs internet on first run; results can be cached and reloaded.

## Libraries

```{r setup, include=FALSE, message=FALSE}
library(data.table)
library(qs2)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(decoupleR)
library(OmnipathR)
library(AnnotationDbi)
library(org.Mm.eg.db)
library(tidytext)
library(DT)
library(htmltools)
```

## Directories

```{r}
base_dir <- "/nemo/lab/destrooperb/home/shared/zanettc/giulia_proteomics/phosphoproteomics"
run_num  <- "run3"

results_dir <- file.path(base_dir, "results", run_num)
objects_dir <- file.path(base_dir, "data", "processed", run_num)
kinase_dir  <- file.path(results_dir, "Kinase_activity")
dir.create(kinase_dir, recursive = TRUE, showWarnings = FALSE)
```

## Load adjusted phospho results

```{r load}
res_df <- fread(file.path(results_dir, "PTM_adjusted_GroupComparison.csv"))

# Parse site identifier into accession + position(s) (same logic as notebook 03)
res_df[, Accession   := str_extract(Protein, "^[^_;]+")]
res_df[, Sites_in_id := str_extract(Protein, "(?<=_)[STY][0-9]+(?:_[STY][0-9]+)*")]
res_df[, Gene := mapIds(org.Mm.eg.db, keys = Accession,
                        column = "SYMBOL", keytype = "UNIPROT",
                        multiVals = "first")]
res_df[, Gene := ifelse(is.na(Gene) | Gene == "", Accession, Gene)]

# decoupleR needs site IDs of form Gene_S123. Expand multi-site rows.
sites_long <- res_df[!is.na(Sites_in_id),
  .(Site = unlist(strsplit(Sites_in_id, "_"))),
  by = .(Label, Gene, log2FC, pvalue, adj.pvalue, Tvalue)
][, ID := paste0(Gene, "_", Site)]
sites_long <- sites_long[!is.na(Gene)]
cat("Site rows after expansion:", nrow(sites_long), "\n")
```

## Pull mouse kinase–substrate network from OmniPath

```{r omnipath, message=FALSE}
ks_net <- OmnipathR::enzyme_substrate(
  organism = 10090
) %>%
  filter(modification == "phosphorylation") %>%
  mutate(
    target = paste0(substrate_genesymbol, "_", residue_type, residue_offset),
    source = enzyme_genesymbol,
    mor    = 1     # all phospho-events are activating substrate phosphorylation
  ) %>%
  dplyr::select(source, target, mor) %>%
  distinct()

cat("Kinase–substrate edges:", nrow(ks_net), "\n")
cat("Unique kinases:        ", length(unique(ks_net$source)), "\n")
qs_save(ks_net, file.path(objects_dir, "omnipath_kinase_substrate.qs2"))
```

## Score per contrast

decoupleR expects a numeric matrix of statistics with sites as rows and one column per "sample" (here: per contrast). We use the t-statistic from MSstatsPTM (`Tvalue`) so direction and magnitude are encoded.

```{r score, message=FALSE}
mat <- sites_long %>%
  dplyr::filter(!is.na(Tvalue)) %>%   # drop NA Tvalue rows — max() returns -Inf on all-NA groups
  group_by(ID, Label) %>%
  summarise(stat = max(Tvalue, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(id_cols = ID, names_from = Label, values_from = stat,
              values_fill = 0) %>%
  as.data.frame()
rownames(mat) <- mat$ID
mat$ID <- NULL
mat <- as.matrix(mat)

# Filter to sites in the OmniPath network — others contribute no signal anyway
mat <- mat[rownames(mat) %in% ks_net$target, , drop = FALSE]
mat <- mat[apply(mat, 1, function(x) all(is.finite(x))), , drop = FALSE]
cat("Sites mapped to OmniPath network:", nrow(mat), "\n")

set.seed(42)
acts_wmean <- decoupleR::run_wmean(mat = mat, network = ks_net,
                                   .source = "source", .target = "target",
                                   .mor = "mor", times = 1000, minsize = 5)
acts_ulm   <- decoupleR::run_ulm(mat = mat, network = ks_net,
                                 .source = "source", .target = "target",
                                 .mor = "mor", minsize = 5)

qs_save(list(wmean = acts_wmean, ulm = acts_ulm),
        file.path(objects_dir, "kinase_activities.qs2"))
```

## Tidy + export per-contrast tables

```{r export}
#Per contrast comparison
tidy <- function(df, method_label) {
  df %>%
    dplyr::filter(statistic == method_label) %>%
    dplyr::group_by(condition) %>%                          # <-- add this
    dplyr::mutate(padj = p.adjust(p_value, method = "BH")) %>%
    dplyr::ungroup() %>%
    dplyr::rename(Kinase = source, Contrast = condition,
                  Activity = score) %>%
    dplyr::select(Contrast, Kinase, Activity, p_value, padj)
}

wmean_tidy <- tidy(acts_wmean, "norm_wmean")
ulm_tidy   <- tidy(acts_ulm,   "ulm")

fwrite(wmean_tidy, file.path(kinase_dir, "Kinase_activity_wmean.csv"))
fwrite(ulm_tidy,   file.path(kinase_dir, "Kinase_activity_ulm.csv"))

sig_kinases <- ulm_tidy %>%
  dplyr::filter(padj < 0.05) %>%
  dplyr::arrange(Contrast, padj)

#Prints padj values in brackets
sig_kinases %>%
  dplyr::arrange(Contrast, padj) %>%
  dplyr::group_by(Contrast) %>%
  dplyr::summarise(
    n      = dplyr::n(),
    kinases = paste0(Kinase, " (", signif(padj, 2), ")", collapse = ", "),
    .groups = "drop"
  ) %>%
  print()
```

## Top kinase barplots per contrast (ULM)
Just top 15 activity scores, with no statistical threshold,
```{r barplots, fig.width=11, fig.height=6}
top_kinases <- ulm_tidy %>%
  group_by(Contrast) %>%
  arrange(desc(abs(Activity))) %>%
  slice_head(n = 15) %>%
  ungroup() %>%
  mutate(Direction = ifelse(Activity > 0, "Activated", "Inhibited"))


p_kin <- ggplot(top_kinases,
                aes(x = reorder_within(Kinase, Activity, Contrast),
                    y = Activity,
                    fill = Direction)) +
  geom_col() +
  scale_x_reordered() +
  scale_fill_manual(values = c("Activated" = "firebrick3",
                               "Inhibited" = "navy")) +
  facet_wrap(~ Contrast, ncol = 3, scales = "free_y") +
  coord_flip() +
  labs(title = "Top 15 kinases by inferred activity per contrast (ULM)",
       x = NULL, y = "Activity score") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top",
        strip.text = element_text(face = "bold"),
        panel.grid.major.y = element_blank())

print(p_kin)
ggsave(file.path(kinase_dir, "Top15_kinases_per_contrast.pdf"),
       p_kin, width = 11, height = 6, device = cairo_pdf)
```

## Substrates of significant kinases (p < 0.05)

For every kinase with padj < 0.05 (ULM), the table below shows up to the top 10 phosphorylated substrates — sites present in the OmniPath network and with nominal p-value < 0.05 in the MSstatsPTM output for that contrast, sorted by p-value.

```{r substrates, results='asis', message=FALSE}
contrasts_ordered <- unique(sig_kinases$Contrast)

for (contrast in contrasts_ordered) {

  kin_rows <- sig_kinases %>% dplyr::filter(Contrast == contrast) %>%
    dplyr::arrange(padj)

  # collect tables first so we can skip contrasts with no substrate data
  kin_tables <- lapply(seq_len(nrow(kin_rows)), function(i) {
    kinase        <- kin_rows$Kinase[i]
    substrate_ids <- ks_net$target[ks_net$source == kinase]

    tbl <- as.data.table(sites_long)[
      Label == contrast & ID %in% substrate_ids & pvalue < 0.05
    ] %>%
      as.data.frame() %>%
      dplyr::arrange(pvalue) %>%
      dplyr::slice_head(n = 10) %>%
      dplyr::mutate(
        log2FC     = round(log2FC,     3),
        Tvalue     = round(Tvalue,     3),
        pvalue     = signif(pvalue,     3),
        adj.pvalue = signif(adj.pvalue, 3)
      ) %>%
      dplyr::select(
        `Substrate site` = ID,
        Gene,
        Site,
        log2FC,
        `T-value`        = Tvalue,
        `p-value`        = pvalue,
        `adj. p-value`   = adj.pvalue
      )
    list(kinase = kinase, padj = kin_rows$padj[i],
         activity = kin_rows$Activity[i], tbl = tbl)
  })

  # drop kinases that have no significant substrates
  kin_tables <- Filter(function(x) nrow(x$tbl) > 0, kin_tables)
  if (length(kin_tables) == 0) next

  cat("### ", contrast, "\n\n")
  cat("::: {.panel-tabset}\n\n")

  for (kt in kin_tables) {
    direction <- ifelse(kt$activity > 0, "Activated", "Inhibited")
    cat("#### ", kt$kinase, "\n\n")
    cat(sprintf("**%s** · activity score %.2f · kinase padj %.3g\n\n",
                direction, kt$activity, kt$padj))

    print(htmltools::tagList(
      DT::datatable(
        kt$tbl,
        rownames   = FALSE,
        filter     = "top",
        extensions = "Buttons",
        options    = list(
          dom        = "Bfrtip",
          buttons    = list("csv", "excel"),
          pageLength = 10,
          scrollX    = TRUE
        )
      )
    ))
    cat("\n\n")
  }

  cat(":::\n\n")
}
```
