---
title: "5. Phospho — Pathway enrichment"
---

## Overview

GSEA on the adjusted phospho contrast results, rolled up to the gene level. For each contrast, the per-gene rank is the most extreme adjusted phospho log2FC across that gene's sites — i.e. the strongest phospho-regulated event the gene exhibits in the contrast. This gives a "phospho-driven pathway" view that complements the bulk GSEA in `bulk_proteomics/notebooks/04_gsea.qmd`.

Same gene-set databases as bulk: GO:BP + KEGG (medicus) and the C8 cell-type signatures, all mouse.

## Libraries

```{r setup, include=FALSE, message=FALSE}
library(data.table)
library(qs2)
library(dplyr)
library(stringr)
library(glue)
library(ggplot2)
library(msigdbr)
library(fgsea)
library(viridis)
library(AnnotationDbi)
library(org.Mm.eg.db)
library(purrr)
```

## Directories

```{r}
base_dir <- "/nemo/lab/destrooperb/home/shared/zanettc/giulia_proteomics/phosphoproteomics"
run_num  <- "run3"

results_dir <- file.path(base_dir, "results", run_num)
objects_dir <- file.path(base_dir, "data", "processed", run_num)
gsea_dir    <- file.path(results_dir, "Phospho_GSEA")
dir.create(gsea_dir, recursive = TRUE, showWarnings = FALSE)
```

## Load adjusted phospho results

```{r load}
res_df <- fread(file.path(results_dir, "PTM_adjusted_GroupComparison.csv"))

res_df[, Accession := str_extract(Protein, "^[^_;]+")]
res_df[, Gene := mapIds(org.Mm.eg.db, keys = Accession,
                        column = "SYMBOL", keytype = "UNIPROT",
                        multiVals = "first")]
res_df[, Gene := ifelse(is.na(Gene) | Gene == "", Accession, Gene)]
```

## Build per-contrast gene-level rank vectors

For each gene, take the site with the largest |adjusted log2FC| and use its log2FC as the gene's rank value.

```{r ranks}
prep_phospho_ranks <- function(df, contrast_label) {
  d <- df %>%
    filter(Label == contrast_label,
           is.finite(log2FC), !is.na(Gene), Gene != "") %>%
    mutate(Gene = toupper(Gene)) %>%
    group_by(Gene) %>%
    slice_max(order_by = abs(log2FC), n = 1, with_ties = FALSE) %>%
    ungroup()
  ranks <- d$log2FC
  names(ranks) <- d$Gene
  ranks <- ranks[!is.na(ranks) & !is.na(names(ranks))]
  sort(ranks, decreasing = TRUE)
}

contrasts_master <- list(
  "hAppNLGF_vs_hApp"            = prep_phospho_ranks(res_df, "hAppNLGF_vs_hApp"),
  "hApp_FIRE_vs_hApp"           = prep_phospho_ranks(res_df, "hApp_FIRE_vs_hApp"),
  "hAppNLGF_FIRE_vs_hAppNLGF"   = prep_phospho_ranks(res_df, "hAppNLGF_FIRE_vs_hAppNLGF"),
  "hAppNLGF_FIRE_vs_hApp_FIRE"  = prep_phospho_ranks(res_df, "hAppNLGF_FIRE_vs_hApp_FIRE"),
  "Interaction_APP_x_Microglia" = prep_phospho_ranks(res_df, "Interaction_APP_x_Microglia")
)
```

## Load mouse gene sets

```{r genesets}
go_df    <- msigdbr(species = "Mus musculus", collection = "C5", subcollection = "GO:BP")
kegg_df  <- msigdbr(species = "Mus musculus", collection = "C2", subcollection = "CP:KEGG_MEDICUS")
c8_df    <- msigdbr(species = "Mus musculus", collection = "C8")

go_list   <- split(go_df$gene_symbol,   go_df$gs_name)
kegg_list <- split(kegg_df$gene_symbol, kegg_df$gs_name)
c8_list   <- split(c8_df$gene_symbol,   c8_df$gs_name)

gene_sets_master <- list(
  "GO_KEGG"           = c(go_list, kegg_list),
  "cell_type_sigs_c8" = c8_list
)
```

## fgsea + barplots (mirrors bulk notebook 04)

```{r run_gsea, message=FALSE, warning=FALSE, fig.width=16, fig.height=10}
run_gsea <- function(ranks, comp_name, geneset_name, pathways) {
  pathways_upper <- lapply(pathways, toupper)
  set.seed(42)
  fgsea_res <- fgsea(pathways    = pathways_upper,
                     stats       = ranks,
                     minSize     = 15,
                     maxSize     = 3000,
                     nPermSimple = 10000,
                     BPPARAM     = BiocParallel::MulticoreParam(workers = 4, progressbar = FALSE))
  if (nrow(fgsea_res) == 0) return(NULL)

  out <- fgsea_res %>%
    as_tibble() %>%
    mutate(leadingEdge = sapply(leadingEdge, paste, collapse = ",")) %>%
    arrange(padj)

  out_dir <- file.path(gsea_dir, comp_name)
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  write.csv(out, file.path(out_dir, glue("Phospho_GSEA_{geneset_name}.csv")),
            row.names = FALSE)

  top <- out %>%
    arrange(desc(NES)) %>% slice_head(n = 10) %>%
    bind_rows(out %>% arrange(NES) %>% slice_head(n = 10)) %>%
    distinct(pathway, .keep_all = TRUE) %>%
    mutate(sig_label = case_when(padj < 0.001 ~ "***",
                                  padj < 0.01  ~ "**",
                                  padj < 0.05  ~ "*",
                                  TRUE         ~ ""),
           pathway = factor(pathway,
                            levels = unique(pathway[order(NES)])))

  if (nrow(top) == 0) return(out)

  axis_zoom <- max(c(2.5, ceiling(max(abs(top$NES)) * 1.05)))
  comp_parts <- str_split(comp_name, "_vs_")[[1]]
  if (length(comp_parts) == 2) {
    label_A <- glue("{comp_parts[1]} Enriched")
    label_B <- glue("{comp_parts[2]} Enriched")
  } else {
    label_A <- "Numerator Enriched"
    label_B <- "Denominator Enriched"
  }

  p <- ggplot(top, aes(x = NES, y = pathway,
                       fill = ifelse(NES > 0, label_A, label_B))) +
    geom_col(width = 0.7) +
    geom_text(aes(label = sig_label,
                  hjust = ifelse(NES > 0, 1.2, -0.2)),
              color = "white", size = 5, vjust = 0.75, fontface = "bold") +
    scale_fill_manual(values = setNames(c("firebrick3", "navy"),
                                         c(label_A, label_B))) +
    labs(x = "Normalized Enrichment Score (NES)", y = NULL,
         title = glue("Phospho GSEA — {gsub('_', ' ', comp_name)} ({geneset_name})")) +
    theme_bw(base_size = 14) +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5, face = "bold", size = 14,
                                     margin = margin(b = 20)),
          panel.grid.major.y = element_blank()) +
    scale_y_discrete(labels = function(x) str_wrap(gsub("_", " ", x), width = 50)) +
    coord_cartesian(xlim = c(-axis_zoom, axis_zoom), clip = "off")

  ggsave(file.path(out_dir, glue("Phospho_GSEA_Barplot_{geneset_name}.pdf")),
         p, width = 9, height = 7)
  print(p)
  out
}

walk2(names(gene_sets_master), gene_sets_master, function(gset_name, pathways) {
  walk2(contrasts_master, names(contrasts_master), function(ranks, comp_name) {
    run_gsea(ranks, comp_name, gset_name, pathways)
    gc()
  })
})
```
