---
title: "6b. Phospho — Site intensity profiles"
---

## Overview

Per-site intensity profiles for all significant adjusted phospho sites (FDR < 0.05, |log2FC| > 0.5) in each contrast. Each panel shows the MSstatsPTM-summarised log₂ intensity across the four conditions, with individual replicate points and a mean ± SE crossbar. Panels are ordered by adjusted p-value (most significant first) and paginated at 20 sites per page.

**Inputs:**

- `PTM_adjusted_GroupComparison.csv` — adjusted DE results from notebook 02
- `summarised_ptm.qs2` → `$PTM$ProteinLevelData` — per-run summarised log₂ intensities

**Thresholds** match notebook 03 (volcano): FDR < 0.05, |log2FC| > 0.5.

## Libraries

```{r setup, include=FALSE, message=FALSE}
library(data.table)
library(qs2)
library(dplyr)
library(stringr)
library(glue)
library(ggplot2)
library(AnnotationDbi)
library(org.Mm.eg.db)
```

## Directories

```{r}
base_dir <- "/nemo/lab/destrooperb/home/shared/zanettc/giulia_proteomics/phosphoproteomics"
run_num  <- "run3"

results_dir <- file.path(base_dir, "results", run_num)
objects_dir <- file.path(base_dir, "data", "processed", run_num)
```

## Condition palette and order

A 2×2 factorial colour scheme: blue family = hApp genotype, red/orange family = NLGF;
dark = microglia present (WT FIRE locus), light = microglia depleted (FIRE KO).

```{r palette}
COND_ORDER <- c("hApp", "hAppNLGF", "hApp_FIRE", "hAppNLGF_FIRE")

cond_cols <- c(
  hApp          = "#2166AC",
  hAppNLGF      = "#D6604D",
  hApp_FIRE     = "#74ADD1",
  hAppNLGF_FIRE = "#F4A582"
)

cond_labels <- c(
  hApp          = "hApp",
  hAppNLGF      = "hApp\nNLGF",
  hApp_FIRE     = "hApp\nFIRE",
  hAppNLGF_FIRE = "hApp\nNLGF\nFIRE"
)
```

## Load adjusted DE results

```{r load_de}
adj <- fread(file.path(results_dir, "PTM_adjusted_GroupComparison.csv"))

adj[, Accession   := str_extract(Protein, "^[^_;]+")]
adj[, Sites_in_id := str_extract(Protein, "(?<=_)[STY][0-9]+(?:_[STY][0-9]+)*")]
adj[, Gene := mapIds(org.Mm.eg.db, keys = Accession,
                     column = "SYMBOL", keytype = "UNIPROT",
                     multiVals = "first")]
adj[, Gene      := ifelse(is.na(Gene) | Gene == "", Accession, Gene)]
adj[, SiteLabel := ifelse(is.na(Sites_in_id), Gene,
                           paste0(Gene, "_", Sites_in_id))]

ADJ_PVAL_CUTOFF <- 0.05
LOGFC_CUTOFF    <- 0.5

adj_sig <- adj[adj.pvalue < ADJ_PVAL_CUTOFF & abs(log2FC) > LOGFC_CUTOFF]
cat("Significant sites per contrast:\n")
print(adj_sig[, .N, by = Label][order(Label)])
```

## Load summarised intensities

```{r load_pld}
summ <- qs_read(file.path(objects_dir, "summarised_ptm.qs2"))
pld  <- as.data.table(summ$PTM$ProteinLevelData)
pld[, GROUP := factor(GROUP, levels = COND_ORDER)]
cat("ProteinLevelData rows:", nrow(pld),
    " | unique sites:", uniqueN(pld$Protein), "\n")
```

## Profile-plot helper

Produces a multi-page `cairo_pdf` for one contrast. Each panel = one significant site,
labelled with gene symbol, residue position, adjusted p-value, and log2FC from the
adjusted DE model. Panels are ordered by ascending adj.pvalue; pages hold ≤ 20 panels.

```{r profile_fn}
PAGE_SIZE <- 20

make_profiles <- function(contrast_label) {
  sig_ct <- adj_sig[Label == contrast_label][order(adj.pvalue)]
  n_sig  <- nrow(sig_ct)
  if (n_sig == 0) {
    message("No significant sites for ", contrast_label, " — skipping.")
    return(invisible(NULL))
  }

  # Build per-panel facet label: Gene_Sxxx\npadj=x.xxx | FC=±x.xx
  sig_ct[, FacetLabel := sprintf(
    "%s\npadj=%.3g | FC=%.2f", SiteLabel, adj.pvalue, log2FC
  )]

  # Ordered factor levels (most significant first)
  sig_ct[, FacetLabel := factor(FacetLabel, levels = unique(FacetLabel))]

  # Pull ProteinLevelData for these sites and join facet labels
  pld_sig <- pld[Protein %in% sig_ct$Protein]
  pld_sig <- merge(pld_sig,
                   sig_ct[, .(Protein, FacetLabel, adj.pvalue)],
                   by = "Protein")
  pld_sig[, FacetLabel := factor(FacetLabel, levels = levels(sig_ct$FacetLabel))]

  n_pages  <- ceiling(n_sig / PAGE_SIZE)
  safe     <- gsub(" ", "_", contrast_label)
  pdf_path <- file.path(results_dir, glue("PhosphoProfiles_{safe}.pdf"))

  plots <- vector("list", n_pages)

  cairo_pdf(pdf_path, onefile = TRUE, width = 14, height = 11)
  for (pg in seq_len(n_pages)) {
    idx_lo   <- (pg - 1L) * PAGE_SIZE + 1L
    idx_hi   <- min(pg * PAGE_SIZE, n_sig)
    pg_sites <- levels(sig_ct$FacetLabel)[idx_lo:idx_hi]
    pg_data  <- pld_sig[FacetLabel %in% pg_sites]
    pg_data[, FacetLabel := factor(FacetLabel, levels = pg_sites)]

    ncol_n <- min(5L, length(pg_sites))
    page_caption <- if (n_pages > 1) glue("Page {pg} of {n_pages}") else NULL

    p <- ggplot(pg_data,
                aes(x = GROUP, y = LogIntensities, colour = GROUP)) +
      stat_summary(fun.data = mean_se,
                   geom     = "crossbar",
                   mapping  = aes(fill = GROUP),
                   width    = 0.55,
                   linewidth = 0.45,
                   fatten   = 1,
                   alpha    = 0.25,
                   show.legend = FALSE) +
      geom_jitter(width = 0.12, size = 1.8, alpha = 0.9) +
      scale_colour_manual(values = cond_cols, limits = COND_ORDER) +
      scale_fill_manual(values   = cond_cols, limits = COND_ORDER) +
      scale_x_discrete(limits = COND_ORDER, labels = cond_labels) +
      facet_wrap(~ FacetLabel, ncol = ncol_n, scales = "free_y") +
      labs(
        x       = NULL,
        y       = "Summarised log₂ intensity",
        title   = glue("Significant adjusted phospho sites — {gsub('_', ' ', contrast_label)}"),
        subtitle = glue(
          "adj.p < {ADJ_PVAL_CUTOFF}  |  |log₂FC| > {LOGFC_CUTOFF}  ",
          "({n_sig} site{ifelse(n_sig==1,'','s')}, phospho minus protein)"
        ),
        caption = page_caption,
        colour  = "Condition"
      ) +
      theme_minimal(base_size = 10) +
      theme(
        plot.title    = element_text(face = "bold", size = 13, hjust = 0),
        plot.subtitle = element_text(size = 9, colour = "grey40",
                                     margin = margin(b = 8)),
        strip.text    = element_text(face = "bold", size = 7.5,
                                     lineheight = 1.15),
        axis.text.x   = element_text(size = 7, lineheight = 0.9),
        axis.text.y   = element_text(size = 7),
        axis.title.y  = element_text(size = 9),
        panel.grid.minor     = element_blank(),
        panel.grid.major.x   = element_blank(),
        panel.border         = element_rect(colour = "grey85", fill = NA,
                                             linewidth = 0.4),
        legend.position      = "top",
        legend.title         = element_text(size = 9, face = "bold"),
        legend.text          = element_text(size = 8)
      ) +
      guides(colour = guide_legend(nrow = 1, override.aes = list(size = 3)))

    plots[[pg]] <- p
    print(p)  # → PDF device
  }
  dev.off()  # close PDF before printing to screen

  for (p in plots) print(p)

  cat("Saved:", pdf_path, "(", n_pages, "page(s),", n_sig, "sites)\n")
  invisible(pdf_path)
}
```

## Generate profile plots — all contrasts

```{r run_profiles, fig.width=14, fig.height=11, message=FALSE}
contrasts_all <- sort(unique(adj_sig$Label))
for (ct in contrasts_all) make_profiles(ct)
```
