---
title: "7. Power downsampling — Interaction contrast"
---

## Overview

Empirical power curve for kinase activity inference on the APP × Microglia interaction contrast, built by subsampling the existing 4-replicates-per-condition data down to N ∈ {2, 3} reps per condition and re-running the full MSstatsPTM → decoupleR pipeline. Each subsample size is repeated `K_DRAWS` times with different random draws; N = 4 is the single reference point.

**What this answers:** "On the data I already have, how steeply does the number of significant kinases drop when I have fewer biological replicates per condition?" A steep drop from N=4→3→2 means more samples will help substantially. A flat curve means N=4 is already near saturation for this contrast (with this site coverage).

**What this does NOT answer:** how many *new* phospho sites would be detected with more samples. More mice typically rescue sites currently lost to feature drop-out — this analysis cannot model that, so its conclusions are a **lower bound** on the gain from increasing N. The site set used at every N here is whatever passes MSstatsPTM's filters on that subset; sites that disappear at lower N reappear at N=4, but we can't conjure sites that don't exist at N=4.

**Inputs (from run3 of notebook 02 + 04):**

- `ptm_input.qs2` — paired PTM + PROTEIN feature-level input (post-localisation + Q-value filter)
- `msstats_annotation.csv` — Run / Condition / BioReplicate mapping
- `omnipath_kinase_substrate.qs2` — mouse kinase–substrate network

## Libraries

```{r setup, include=FALSE, message=FALSE}
library(data.table)
library(qs2)
library(MSstats)
library(MSstatsPTM)
library(decoupleR)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(AnnotationDbi)
library(org.Mm.eg.db)
library(foreach)
library(doSNOW)
library(parallel)
library(RhpcBLASctl)
```

## Directories and parameters

```{r}
base_dir <- "/nemo/lab/destrooperb/home/shared/zanettc/giulia_proteomics/phosphoproteomics"
run_num  <- "run3"

results_dir <- file.path(base_dir, "results", run_num)
objects_dir <- file.path(base_dir, "data", "processed", run_num)
power_dir   <- file.path(results_dir, "Power_downsampling")
draws_dir   <- file.path(objects_dir, "power_draws")
dir.create(power_dir,  recursive = TRUE, showWarnings = FALSE)
dir.create(draws_dir,  recursive = TRUE, showWarnings = FALSE)

# Subsample sizes per condition (full = 4). Baseline N=4 is added automatically.
N_VALUES <- c(2L, 3L)
K_DRAWS  <- 10L             # random draws per N
SEED     <- 42L

CONTRAST_NAME <- "Interaction_APP_x_Microglia"

# Significance thresholds — match notebook 03 / 04
SITE_PADJ_CUT <- 0.05
SITE_FC_CUT   <- 0.5
KINASE_PADJ_CUT <- 0.05

# Parallel workers. Defaults to N=21 (the number of draws when K_DRAWS=10:
# 1 + 10 + 10). Override via the SLURM_CPUS_PER_TASK env var so the same
# notebook scales to whatever the SLURM job asks for. Clamped to n_draws.
N_CORES <- as.integer(Sys.getenv("SLURM_CPUS_PER_TASK", "21"))

# Paths needed by each worker. Workers re-load the cached input from disk
# (rather than receiving 1.5+ GB tables via doSNOW serialisation) and
# subset to their assigned runs before doing any work.
PTM_INPUT_PATH <- file.path(objects_dir, "ptm_input.qs2")
KS_NET_PATH    <- file.path(objects_dir, "omnipath_kinase_substrate.qs2")
```

## Load annotation

The full PTM + PROTEIN tables are NOT loaded in the parent process — each worker re-loads from `PTM_INPUT_PATH` and subsets to its kept runs immediately, to keep peak memory bounded.

```{r load}
ann <- fread(file.path(objects_dir, "msstats_annotation.csv"))
cat("Annotation:\n"); print(ann)
```

## Build draw plan

Every draw picks `N` BioReplicates from each of the four conditions (sampling without replacement), seeded for reproducibility. The full set (N = 4) is included once.

```{r draw_plan}
set.seed(SEED)
conds <- unique(ann$Condition)

make_draw <- function(N, seed) {
  set.seed(seed)
  do.call(c, lapply(conds, function(co) {
    runs_co <- ann[Condition == co, Run]
    sample(runs_co, N)
  }))
}

draw_plan <- rbindlist(lapply(N_VALUES, function(N) {
  data.table(
    N        = N,
    draw_id  = seq_len(K_DRAWS),
    seed     = SEED + 1000L * N + seq_len(K_DRAWS),
    runs_csv = vapply(seq_len(K_DRAWS),
                      function(k) paste(make_draw(N, SEED + 1000L * N + k),
                                        collapse = ","),
                      character(1))
  )
}))

# Add the N=4 baseline (no random subset — all 16 samples)
draw_plan <- rbindlist(list(
  data.table(N = 4L, draw_id = 0L, seed = NA_integer_,
             runs_csv = paste(ann$Run, collapse = ",")),
  draw_plan
), use.names = TRUE)

draw_plan[, file := file.path(draws_dir,
                              sprintf("draw_N%d_d%02d.qs2", N, draw_id))]
print(draw_plan[, .(N, draw_id, seed, n_runs = lengths(strsplit(runs_csv, ",")))])
```

## Pipeline function — one (N, draw) → kinase activities

Re-runs `dataSummarizationPTM` → `groupComparisonPTM` (Interaction contrast only) → `decoupleR::run_ulm`. Both PTM and PROTEIN are subsetted to the same kept runs so the protein-level adjustment is consistent.

```{r pipeline_fn}
run_one_draw <- function(kept_runs, ann_full,
                         ptm_input_path, ks_net_path,
                         contrast_name,
                         site_padj_cut, site_fc_cut, kinase_padj_cut) {

  # Pin BLAS/OMP threads to 1 inside the worker — env vars also set
  # this from the SLURM script, but belt-and-braces against thread
  # oversubscription when many workers run MSstatsPTM in parallel.
  try(RhpcBLASctl::blas_set_num_threads(1), silent = TRUE)
  try(RhpcBLASctl::omp_set_num_threads(1),  silent = TRUE)

  # Worker reads the cached input itself so the 1.5+ GB tables are
  # not serialised through the doSNOW socket. Subset immediately, drop
  # the full object, and gc to keep peak memory bounded.
  ptm_input_full <- qs2::qs_read(ptm_input_path)
  ks_net         <- qs2::qs_read(ks_net_path)

  ptm_sub  <- as.data.table(ptm_input_full$PTM    )[Run %in% kept_runs]
  prot_sub <- as.data.table(ptm_input_full$PROTEIN)[Run %in% kept_runs]
  rm(ptm_input_full); gc(verbose = FALSE)

  ann_sub <- ann_full[Run %in% kept_runs]

  ptm_input_sub <- list(
    PTM     = as.data.frame(ptm_sub),
    PROTEIN = as.data.frame(prot_sub)
  )

  summ <- suppressWarnings(dataSummarizationPTM(
    data                                 = ptm_input_sub,
    normalization                        = "equalizeMedians",
    summaryMethod                        = "TMP",
    censoredInt                          = "NA",
    MBimpute                             = TRUE,
    featureSubset                        = "highQuality",
    remove_uninformative_feature_outlier = TRUE
  ))

  cond_levels <- levels(summ$PTM$ProteinLevelData$GROUP)
  stopifnot(all(c("hApp","hAppNLGF","hApp_FIRE","hAppNLGF_FIRE") %in% cond_levels))

  # Interaction-only contrast row
  w <- setNames(rep(0, length(cond_levels)), cond_levels)
  w[c("hApp","hAppNLGF","hApp_FIRE","hAppNLGF_FIRE")] <- c(-1, 1, 1, -1)
  cm <- matrix(w, nrow = 1, dimnames = list(contrast_name, cond_levels))

  res <- suppressWarnings(groupComparisonPTM(
    data            = summ,
    contrast.matrix = cm,
    data.type       = "LabelFree"
  ))

  adj <- as.data.table(res$ADJUSTED.Model)

  # Map site IDs the same way as notebook 04
  adj[, Accession   := str_extract(Protein, "^[^_;]+")]
  adj[, Sites_in_id := str_extract(Protein, "(?<=_)[STY][0-9]+(?:_[STY][0-9]+)*")]
  adj[, Gene := mapIds(org.Mm.eg.db, keys = Accession,
                       column = "SYMBOL", keytype = "UNIPROT",
                       multiVals = "first")]
  adj[, Gene := ifelse(is.na(Gene) | Gene == "", Accession, Gene)]

  sites_long <- adj[!is.na(Sites_in_id) & !is.na(Gene) & !is.na(Tvalue),
    .(Site = unlist(strsplit(Sites_in_id, "_"))),
    by = .(Label, Gene, log2FC, pvalue, adj.pvalue, Tvalue)
  ][, ID := paste0(Gene, "_", Site)]

  # Build the t-stat matrix (one column: this contrast)
  mat <- sites_long[, .(stat = max(Tvalue, na.rm = TRUE)),
                    by = .(ID, Label)] %>%
    tidyr::pivot_wider(id_cols = ID, names_from = Label,
                       values_from = stat, values_fill = 0) %>%
    as.data.frame()
  rownames(mat) <- mat$ID; mat$ID <- NULL
  mat <- as.matrix(mat)
  mat <- mat[rownames(mat) %in% ks_net$target, , drop = FALSE]
  mat <- mat[apply(mat, 1, function(x) all(is.finite(x))), , drop = FALSE]

  acts_ulm <- decoupleR::run_ulm(
    mat = mat, network = ks_net,
    .source = "source", .target = "target",
    .mor = "mor", minsize = 5
  )

  ulm_tidy <- acts_ulm %>%
    dplyr::filter(statistic == "ulm") %>%
    dplyr::mutate(padj = p.adjust(p_value, method = "BH")) %>%
    dplyr::rename(Kinase = source, Activity = score) %>%
    dplyr::select(Kinase, Activity, p_value, padj) %>%
    as.data.table()

  # Summary scalars
  n_sig_sites <- adj[adj.pvalue < site_padj_cut &
                     abs(log2FC) > site_fc_cut, .N]
  sig_kin     <- ulm_tidy[padj < kinase_padj_cut][order(padj)]

  list(
    adj          = adj,
    ulm_tidy     = ulm_tidy,
    n_sig_sites  = n_sig_sites,
    n_sig_kinases = nrow(sig_kin),
    sig_kinases  = sig_kin$Kinase
  )
}
```

## Run all draws in parallel (resumable)

Each draw is cached at `data/processed/run3/power_draws/draw_N{N}_d{draw_id}.qs2`. The block below:

1. Filters `draw_plan` to draws without a cached result.
2. If all draws are cached (e.g. when rendering the Quarto HTML after a SLURM compute job), skips the cluster entirely.
3. Otherwise spins up `min(N_CORES, n_uncached)` `doSNOW` workers, each of which loads `ptm_input.qs2` from disk itself and subsets to its kept runs before running MSstatsPTM. `.errorhandling = "pass"` lets the loop survive a single draw failure (e.g. degenerate N=2 design).

```{r run_loop, message=FALSE}
to_run <- draw_plan[!file.exists(file)]

if (nrow(to_run) == 0) {
  cat("All draws cached — skipping cluster setup.\n")
} else {
  n_workers <- min(N_CORES, nrow(to_run))
  cat(sprintf("Spinning up %d worker(s) for %d uncached draw(s)...\n",
              n_workers, nrow(to_run)))

  # outfile = "" streams worker stdout/stderr to the parent process so
  # SLURM logs capture per-draw progress lines.
  cl <- parallel::makeCluster(n_workers, type = "SOCK", outfile = "")
  doSNOW::registerDoSNOW(cl)
  on.exit(try(parallel::stopCluster(cl), silent = TRUE), add = TRUE)

  WORKER_PACKAGES <- c(
    "data.table", "qs2", "MSstats", "MSstatsPTM", "decoupleR",
    "dplyr", "tidyr", "stringr", "AnnotationDbi", "org.Mm.eg.db",
    "RhpcBLASctl"
  )

  results <- foreach(
    i = seq_len(nrow(to_run)),
    .packages     = WORKER_PACKAGES,
    .export       = c("run_one_draw", "ann", "to_run",
                      "PTM_INPUT_PATH", "KS_NET_PATH",
                      "CONTRAST_NAME",
                      "SITE_PADJ_CUT", "SITE_FC_CUT", "KINASE_PADJ_CUT"),
    .errorhandling = "pass"
  ) %dopar% {

    kept <- strsplit(to_run$runs_csv[i], ",")[[1]]
    t0   <- Sys.time()

    out <- run_one_draw(
      kept_runs       = kept,
      ann_full        = ann,
      ptm_input_path  = PTM_INPUT_PATH,
      ks_net_path     = KS_NET_PATH,
      contrast_name   = CONTRAST_NAME,
      site_padj_cut   = SITE_PADJ_CUT,
      site_fc_cut     = SITE_FC_CUT,
      kinase_padj_cut = KINASE_PADJ_CUT
    )

    out$meta <- list(
      N       = to_run$N[i],
      draw_id = to_run$draw_id[i],
      seed    = to_run$seed[i],
      runs    = kept,
      elapsed = as.numeric(difftime(Sys.time(), t0, units = "secs"))
    )
    qs2::qs_save(out, to_run$file[i])

    cat(sprintf("[worker pid=%d] N=%d draw=%d  →  %d sig sites, %d sig kinases (%.1f min)\n",
                Sys.getpid(),
                to_run$N[i], to_run$draw_id[i],
                out$n_sig_sites, out$n_sig_kinases,
                out$meta$elapsed / 60))

    data.table(
      N             = to_run$N[i],
      draw_id       = to_run$draw_id[i],
      n_sig_sites   = out$n_sig_sites,
      n_sig_kinases = out$n_sig_kinases,
      elapsed_min   = out$meta$elapsed / 60,
      status        = "ok"
    )
  }
  parallel::stopCluster(cl)

  # Report any iterations that came back as error objects (errorhandling=pass).
  bad <- which(vapply(results, inherits, logical(1), what = "error"))
  if (length(bad)) {
    cat(sprintf("\n%d draw(s) failed; their cache files will be missing:\n",
                length(bad)))
    for (k in bad) {
      cat(sprintf("  - draw row %d: %s\n", k, conditionMessage(results[[k]])))
    }
  }
}
```

## Aggregate

```{r aggregate}
all_draws <- rbindlist(lapply(draw_plan$file, function(f) {
  x <- qs_read(f)
  data.table(
    N             = x$meta$N,
    draw_id       = x$meta$draw_id,
    n_sig_sites   = x$n_sig_sites,
    n_sig_kinases = x$n_sig_kinases,
    sig_kinases   = paste(x$sig_kinases, collapse = ", ")
  )
}))
all_draws[, N := factor(N, levels = sort(unique(N)))]
print(all_draws[order(N, draw_id)])

fwrite(all_draws, file.path(power_dir, "power_summary.csv"))
```

## Per-kinase persistence across draws

For each kinase, count how often it reaches FDR < 0.05 across the K draws at each N. Kinases that are robustly significant at low N are the "easy wins"; kinases that appear only at N=4 are the ones losing power from sample size.

```{r kinase_persistence}
per_kin <- rbindlist(lapply(draw_plan$file, function(f) {
  x <- qs_read(f)
  if (length(x$sig_kinases) == 0) {
    return(data.table(N = integer(), draw_id = integer(), Kinase = character()))
  }
  data.table(N = x$meta$N,
             draw_id = x$meta$draw_id,
             Kinase  = x$sig_kinases)
}))

n_per_N <- draw_plan[, .(n_draws = uniqueN(draw_id)), by = N]

if (nrow(per_kin) > 0) {
  persistence <- per_kin[, .(n_hit = .N), by = .(N, Kinase)]
  persistence <- merge(persistence, n_per_N, by = "N")
  persistence[, hit_rate := n_hit / n_draws]
  persistence <- persistence[order(N, -hit_rate, Kinase)]
} else {
  persistence <- data.table(N = integer(), Kinase = character(),
                            n_hit = integer(), n_draws = integer(),
                            hit_rate = numeric())
}
print(persistence)

fwrite(persistence, file.path(power_dir, "kinase_persistence.csv"))
```

## Plot — power curve

```{r plot_power, fig.width=7, fig.height=5}
p <- ggplot(all_draws,
            aes(x = N, y = n_sig_kinases)) +
  geom_jitter(width = 0.12, height = 0, size = 2.5,
              alpha = 0.7, colour = "#2166AC") +
  stat_summary(fun = mean, geom = "crossbar",
               width = 0.4, colour = "grey20", fatten = 1.2) +
  labs(
    title    = "Power curve — Interaction contrast",
    subtitle = sprintf(
      "K=%d random draws per N (N=4 is full data, single point)  •  ULM padj < %.2f",
      K_DRAWS, KINASE_PADJ_CUT),
    x = "Biological replicates per condition",
    y = "Significant kinases (FDR < 0.05)"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())

print(p)
ggsave(file.path(power_dir, "power_curve_kinases.pdf"),
       p, width = 7, height = 5, device = cairo_pdf)

p_sites <- ggplot(all_draws,
                  aes(x = N, y = n_sig_sites)) +
  geom_jitter(width = 0.12, height = 0, size = 2.5,
              alpha = 0.7, colour = "#D6604D") +
  stat_summary(fun = mean, geom = "crossbar",
               width = 0.4, colour = "grey20", fatten = 1.2) +
  labs(
    title    = "Power curve — sig phospho sites (Interaction)",
    subtitle = sprintf("adj.p < %.2f  &  |log2FC| > %.2f", SITE_PADJ_CUT, SITE_FC_CUT),
    x = "Biological replicates per condition",
    y = "Significant sites"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())

print(p_sites)
ggsave(file.path(power_dir, "power_curve_sites.pdf"),
       p_sites, width = 7, height = 5, device = cairo_pdf)
```

## Interpretation cues

- **Mean curve still rising at N=4** → more mice will likely raise the kinase count. Magnitude of expected gain is bounded by the slope here (and is in fact larger, because more N also detects new sites).
- **Mean curve plateauing at N=4** → either the contrast is genuinely null beyond a small set of strong kinases, or the bottleneck is site coverage (not N). In that case, switching to a deeper acquisition or relaxing site-level filters may help more than adding mice.
- **High variance across draws at N=2 or N=3** → indicates fragile inference. Look at `kinase_persistence.csv` to see which kinases come and go between draws vs. which are stable.
