---
title: "6. PPI Network Analysis"
format: html
---

## Overview

Protein–protein interaction networks for the significant proteins of each MSstats contrast, using STRINGdb (mouse, physical interactions, score ≥ 700). For each contrast we build separate networks for upregulated and downregulated proteins, identify hub proteins by degree centrality, and export per-contrast hub tables.

Adapted from `millie_proteomics/notebooks/08_PPI.qmd`, with this project's contrast set (four pairwise + the APP × Microglia interaction).

## Libraries

```{r setup, message=TRUE, warning=FALSE,}
library(data.table)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
library(stringr)
library(STRINGdb)
library(igraph)
library(qs2)
library(scales)
```

## Directories and parameters

```{r}
base_dir    <- "/nemo/lab/destrooperb/home/shared/zanettc/giulia_proteomics/bulk_proteomics"
run_num     <- "run3"
results_dir <- file.path(base_dir, "results", run_num)
objects_dir <- file.path(base_dir, "data", "processed", run_num)
ppi_dir     <- file.path(results_dir, "PPI_networks")
dir.create(ppi_dir, recursive = TRUE, showWarnings = FALSE)

FDR_CUTOFF   <- 0.05
LOGFC_CUTOFF <- 0.5
STRING_SCORE <- 700
TOP_N        <- 30   # nodes shown in the top-N subnetwork plot
```

## Load and prepare DE results

```{r}
res_df             <- fread(file.path(results_dir, "Full_GroupComparison_Results.csv"))
protein_dictionary <- fread(file.path(objects_dir, "protein_dictionary.csv"))

msstats_ready <- res_df %>%
  filter(!is.na(adj.pvalue), is.finite(log2FC)) %>%
  left_join(protein_dictionary, by = "Protein") %>%
  mutate(
    Gene = ifelse(is.na(Gene) | Gene == "", Protein, Gene),
    Status = case_when(
      adj.pvalue < FDR_CUTOFF & log2FC >  LOGFC_CUTOFF ~ "Upregulated",
      adj.pvalue < FDR_CUTOFF & log2FC < -LOGFC_CUTOFF ~ "Downregulated",
      TRUE ~ "Not Significant"
    )
  )
```

## Initialise STRINGdb (mouse, physical, high-confidence)

```{r}
# Crick cluster blocks HTTPS with SSL; wget --no-check-certificate bypasses it
options(download.file.method = "wget",
        download.file.extra  = "--no-check-certificate")

stringdb_cache <- "/nemo/lab/destrooperb/home/shared/zanettc/giulia_proteomics/bulk_proteomics/data/stringdb_cache"
dir.create(stringdb_cache, recursive = TRUE, showWarnings = FALSE)

string_db <- STRINGdb$new(
  version         = "12.0",
  species         = 10090,
  score_threshold = STRING_SCORE,
  network_type    = "physical",
  input_directory = stringdb_cache
)
```

## Build networks per contrast and direction

```{r}
contrasts <- c(
  "hAppNLGF_vs_hApp",
  "hApp_FIRE_vs_hApp",
  "hAppNLGF_FIRE_vs_hAppNLGF",
  "hAppNLGF_FIRE_vs_hApp_FIRE",
  "Interaction_APP_x_Microglia"
)

build_network <- function(genes_df) {
  if (nrow(genes_df) < 5) return(NULL)
  mapped <- string_db$map(genes_df, "Gene", removeUnmappedRows = TRUE)
  if (nrow(mapped) < 5) return(NULL)
  inter <- string_db$get_interactions(mapped$STRING_id)
  if (nrow(inter) == 0)
    return(list(proteins = mapped, interactions = inter, graph = NULL))
  g <- graph_from_data_frame(
    inter %>% dplyr::select(from, to, combined_score),
    directed = FALSE,
    vertices = mapped %>% dplyr::select(STRING_id, Gene, log2FC)
  ) %>% simplify()
  V(g)$degree <- degree(g)
  list(proteins = mapped, interactions = inter, graph = g)
}

network_results <- list()

for (comp in contrasts) {
  cat("\n============================\n", comp, "\n============================\n")
  for (dir_lab in c("Upregulated", "Downregulated")) {
    sig <- msstats_ready %>%
      filter(Label == comp, Status == dir_lab) %>%
      dplyr::select(Gene, log2FC, adj.pvalue) %>%
      distinct(Gene, .keep_all = TRUE) %>%
      as.data.frame()
    cat(sprintf("  %-13s n = %d\n", dir_lab, nrow(sig)))
    nr <- build_network(sig)
    if (!is.null(nr) && !is.null(nr$graph))
      cat(sprintf("    -> nodes %d | edges %d\n",
                  vcount(nr$graph), ecount(nr$graph)))
    network_results[[paste(comp, dir_lab, sep = "__")]] <- nr
  }
}
```

## Visualise full networks (one PDF per contrast/direction)

```{r plot_full, fig.width=14, fig.height=14}
plot_full_network <- function(nr, label) {
  if (is.null(nr) || is.null(nr$graph) || ecount(nr$graph) == 0) return(invisible())
  g_plot <- delete_vertices(nr$graph, V(nr$graph)[degree(nr$graph) == 0])
  if (vcount(g_plot) < 3) return(invisible())

  V(g_plot)$label       <- V(g_plot)$Gene
  V(g_plot)$size        <- rescale(degree(g_plot), to = c(5, 20))
  V(g_plot)$label.cex   <- rescale(degree(g_plot), to = c(0.6, 1.3))
  V(g_plot)$label.font  <- 2
  V(g_plot)$color       <- "steelblue"
  V(g_plot)$frame.color <- "grey30"
  E(g_plot)$color       <- "grey75"
  E(g_plot)$width       <- 0.8

  top_hubs <- names(sort(degree(g_plot), decreasing = TRUE))[1:min(10, vcount(g_plot))]
  V(g_plot)$color[V(g_plot)$name %in% top_hubs] <- "firebrick3"

  pdf(file.path(ppi_dir, paste0("network_full_", label, ".pdf")), width = 14, height = 14)
  set.seed(42)
  plot(g_plot,
       layout = layout_with_fr(g_plot),
       vertex.label.color = "black",
       main = paste0(label, "\nSTRING physical >= ", STRING_SCORE,
                     " | Nodes: ", vcount(g_plot),
                     " | Edges: ", ecount(g_plot),
                     " | Red = top 10 hubs"))
  dev.off()

  set.seed(42)
  plot(g_plot,
       layout = layout_with_fr(g_plot),
       vertex.label.color = "black",
       main = paste0(label, "\nNodes: ", vcount(g_plot),
                     " | Edges: ", ecount(g_plot)))
}

for (key in names(network_results))
  plot_full_network(network_results[[key]], key)
```

## Top-N subnetwork plots — focused view of strongest hubs

```{r plot_topN, fig.width=12, fig.height=12}
plot_top_network <- function(nr, label, top_n = TOP_N) {
  if (is.null(nr) || is.null(nr$graph) || ecount(nr$graph) == 0) return(invisible())
  g <- nr$graph
  top_nodes <- names(sort(degree(g), decreasing = TRUE))[1:min(top_n, vcount(g))]
  g_top <- induced_subgraph(g, top_nodes) %>%
    (function(x) delete_vertices(x, V(x)[degree(x) == 0]))()
  if (vcount(g_top) < 3) return(invisible())

  V(g_top)$label       <- V(g_top)$Gene
  V(g_top)$size        <- rescale(degree(g_top), to = c(8, 22))
  V(g_top)$label.cex   <- 1.0
  V(g_top)$label.font  <- 2
  V(g_top)$frame.color <- "grey30"
  E(g_top)$color       <- "grey70"
  E(g_top)$width       <- 1.2

  fc <- V(g_top)$log2FC
  pal <- colorRampPalette(c("navy", "white", "firebrick3"))(100)
  fc_idx <- findInterval(fc, seq(min(fc, na.rm = TRUE),
                                 max(fc, na.rm = TRUE),
                                 length.out = 100))
  fc_idx[fc_idx == 0] <- 1
  V(g_top)$color <- pal[fc_idx]

  pdf(file.path(ppi_dir, paste0("network_top", top_n, "_", label, ".pdf")),
      width = 12, height = 12)
  set.seed(42)
  plot(g_top, layout = layout_with_fr(g_top),
       vertex.label.color = "black",
       main = paste0(label, " — top ", vcount(g_top), " by degree\n",
                     "Coloured by log2FC | Edges: ", ecount(g_top)))
  legend_vals   <- seq(min(fc, na.rm = TRUE), max(fc, na.rm = TRUE), length.out = 5)
  legend_colors <- colorRampPalette(c("navy", "white", "firebrick3"))(5)
  legend("bottomright", legend = round(legend_vals, 2),
         fill = legend_colors, title = "log2FC", bty = "n", cex = 0.8)
  dev.off()

  set.seed(42)
  plot(g_top, layout = layout_with_fr(g_top),
       vertex.label.color = "black",
       main = paste0(label, " — top ", vcount(g_top), " by degree"))
  legend("bottomright", legend = round(legend_vals, 2),
         fill = legend_colors, title = "log2FC", bty = "n", cex = 0.8)
}

for (key in names(network_results))
  plot_top_network(network_results[[key]], key)
```

## Hub protein summary tables

```{r}
hub_list <- list()

for (key in names(network_results)) {
  nr <- network_results[[key]]
  if (is.null(nr) || is.null(nr$graph)) next
  g <- nr$graph
  hub_df <- data.frame(
    Gene       = V(g)$Gene,
    log2FC     = V(g)$log2FC,
    Degree     = degree(g),
    Comparison = key
  ) %>% filter(Degree > 0) %>% arrange(desc(Degree))
  hub_list[[key]] <- hub_df
  fwrite(hub_df, file.path(ppi_dir, paste0("hub_proteins_", key, ".csv")))

  cat("\n--- Top 20 hubs:", key, "---\n")
  print(head(hub_df, 20))
}

if (length(hub_list) > 0) {
  hub_combined <- bind_rows(hub_list)
  fwrite(hub_combined, file.path(ppi_dir, "hub_proteins_all.csv"))
}
```

## Network summary statistics

```{r}
summary_table <- data.frame(
  Comparison       = character(),
  Direction        = character(),
  Sig_proteins     = integer(),
  Mapped_to_STRING = integer(),
  Nodes_with_edges = integer(),
  Edges            = integer(),
  stringsAsFactors = FALSE
)

for (key in names(network_results)) {
  parts <- strsplit(key, "__", fixed = TRUE)[[1]]
  comp <- parts[1]; dir_lab <- parts[2]
  nr <- network_results[[key]]
  g  <- if (!is.null(nr)) nr$graph else NULL
  n_with_edges <- if (!is.null(g)) sum(degree(g) > 0) else 0
  n_edges      <- if (!is.null(g)) ecount(g) else 0
  sig_n <- nrow(msstats_ready %>%
                  filter(Label == comp, Status == dir_lab) %>%
                  distinct(Gene))
  summary_table <- rbind(summary_table, data.frame(
    Comparison       = comp,
    Direction        = dir_lab,
    Sig_proteins     = sig_n,
    Mapped_to_STRING = if (!is.null(nr)) nrow(nr$proteins) else 0,
    Nodes_with_edges = n_with_edges,
    Edges            = n_edges
  ))
}

print(summary_table)
fwrite(summary_table, file.path(ppi_dir, "network_summary.csv"))
```

## Session info

```{r}
sessionInfo()
```
