---
title: "7. Subcellular and secretome stratification"
format: html
---

## Overview

Stratifies the significant proteins from each MSstats contrast by subcellular compartment using GO Cellular Component annotations (MSigDB C5 GO:CC). Useful for spotting compartment-skewed responses (e.g. "microglia depletion is dominated by lysosomal/phagosomal proteins") and for flagging potential secreted biomarkers.

## Libraries

```{r setup, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(data.table)
  library(qs2)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(ggplot2)
  library(ggrepel)
  library(msigdbr)
  library(scales)
})
```

## Directories and parameters

```{r}
base_dir    <- "/nemo/lab/destrooperb/home/shared/zanettc/giulia_proteomics/bulk_proteomics"
run_num     <- "run3"
results_dir <- file.path(base_dir, "results", run_num)
objects_dir <- file.path(base_dir, "data", "processed", run_num)

subcell_dir <- file.path(results_dir, "Subcellular")
dir.create(subcell_dir, recursive = TRUE, showWarnings = FALSE)

FDR_CUTOFF   <- 0.05
LOGFC_CUTOFF <- 0.5
```

## Load DE results and protein dictionary

```{r}
res_df             <- fread(file.path(results_dir, "Full_GroupComparison_Results.csv"))
protein_dictionary <- fread(file.path(objects_dir, "protein_dictionary.csv"))

msstats_ready <- res_df %>%
  filter(!is.na(adj.pvalue), is.finite(log2FC)) %>%
  left_join(protein_dictionary, by = "Protein") %>%
  mutate(
    Gene = ifelse(is.na(Gene) | Gene == "", Protein, Gene),
    Gene_upper = toupper(Gene),
    Status = case_when(
      adj.pvalue < FDR_CUTOFF & log2FC >  LOGFC_CUTOFF ~ "Upregulated",
      adj.pvalue < FDR_CUTOFF & log2FC < -LOGFC_CUTOFF ~ "Downregulated",
      TRUE ~ "Not Significant"
    )
  )
```

## Build compartment → genes mapping from GO:CC

We assign each gene to one or more broad compartments based on the GO Cellular Component gene sets (mouse). The compartment categories are defined by keyword matches against GO:CC term names — coarser than raw GO IDs but readable.

```{r}
gocc <- msigdbr(species = "Mus musculus", collection = "C5", subcollection = "GO:CC")

# Compartment definitions — order matters (first match wins on the per-gene roll-up)
compartment_patterns <- list(
  "Extracellular / Secreted" = "EXTRACELLULAR|SECRET",
  "Plasma membrane"          = "PLASMA_MEMBRANE|CELL_SURFACE|CELL_PROJECTION_MEMBRANE",
  "Lysosome / Endosome"      = "LYSOSOM|ENDOSOM|PHAGOSOM|AUTOPHAG",
  "Mitochondrion"            = "MITOCHONDRI",
  "Endoplasmic reticulum"    = "ENDOPLASMIC_RETICULUM|\\bER\\b",
  "Golgi"                    = "GOLGI",
  "Nucleus"                  = "NUCLEAR|NUCLEUS|NUCLEOLUS|NUCLEOPLASM|CHROMATIN",
  "Cytoskeleton"             = "CYTOSKELET|MICROTUBULE|ACTIN_FILAMENT|INTERMEDIATE_FILAMENT",
  "Ribosome / Translation"   = "RIBOSOM|POLYSOM|TRANSLATION_INITIATION",
  "Synapse"                  = "SYNAPS|POSTSYNAP|PRESYNAP|AXON|DENDRITE",
  "Cytosol / Cytoplasm"      = "CYTOSOL|\\bCYTOPLASM\\b"
)

# Long-format gene × compartment table (one gene can hit multiple compartments)
compartment_map <- bind_rows(lapply(names(compartment_patterns), function(cmp) {
  pat <- compartment_patterns[[cmp]]
  gocc %>%
    filter(str_detect(gs_name, pat)) %>%
    distinct(gene_symbol) %>%
    transmute(Compartment = cmp, Gene_upper = toupper(gene_symbol))
})) %>%
  distinct()

cat("Compartment coverage (genes per compartment):\n")
compartment_map %>% count(Compartment, sort = TRUE) %>% print()
```

## Annotate the significant proteins

```{r}
sig <- msstats_ready %>% filter(Status != "Not Significant")

sig_compart <- sig %>%
  inner_join(compartment_map, by = "Gene_upper", relationship = "many-to-many")

# Coverage diagnostic: significant genes that hit no compartment
no_compart <- setdiff(unique(sig$Gene_upper), unique(compartment_map$Gene_upper))
cat("Significant genes with no GO:CC annotation in our compartment list:",
    length(no_compart), "/", length(unique(sig$Gene_upper)), "\n")
```

## Per-compartment counts of significant proteins per contrast

```{r compartment_counts, fig.width=11, fig.height=6}
counts_df <- sig_compart %>%
  count(Label, Compartment, Status) %>%
  mutate(n_signed = ifelse(Status == "Upregulated", n, -n))

p_counts <- ggplot(counts_df, aes(x = reorder(Compartment, abs(n_signed)),
                                  y = n_signed, fill = Status)) +
  geom_col() +
  scale_fill_manual(values = c("Upregulated" = "firebrick3",
                                "Downregulated" = "navy")) +
  facet_wrap(~ Label, ncol = 3, scales = "free_y") +
  coord_flip() +
  labs(x = NULL, y = "# significant proteins (- = down, + = up)",
       title = "Significant proteins by subcellular compartment",
       subtitle = paste0("FDR < ", FDR_CUTOFF, ", |log2FC| > ", LOGFC_CUTOFF)) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top",
        strip.text       = element_text(face = "bold"),
        panel.grid.major.y = element_blank())

print(p_counts)
ggsave(file.path(subcell_dir, "Compartment_counts_by_contrast.pdf"),
       p_counts, width = 11, height = 6, device = cairo_pdf)
```

## Stratified volcano — facet by compartment for one contrast at a time

```{r facet_volcano, fig.width=12, fig.height=10}
plot_compartment_volcano <- function(contrast_label) {
  sub <- msstats_ready %>%
    filter(Label == contrast_label) %>%
    inner_join(compartment_map, by = "Gene_upper", relationship = "many-to-many")

  if (nrow(sub) == 0) return(invisible())

  labels <- sub %>% filter(Status != "Not Significant") %>%
    group_by(Compartment, Status) %>%
    arrange(adj.pvalue) %>%
    slice_head(n = 5) %>%
    ungroup()

  p <- ggplot(sub, aes(x = log2FC, y = -log10(adj.pvalue))) +
    geom_point(aes(color = Status, size = Status, alpha = Status), shape = 16) +
    scale_color_manual(values = c("Upregulated" = "firebrick3",
                                   "Downregulated" = "navy",
                                   "Not Significant" = "grey80")) +
    scale_size_manual(values  = c("Upregulated" = 1.6, "Downregulated" = 1.6,
                                   "Not Significant" = 0.7)) +
    scale_alpha_manual(values = c("Upregulated" = 0.9, "Downregulated" = 0.9,
                                   "Not Significant" = 0.4)) +
    geom_hline(yintercept = -log10(FDR_CUTOFF), linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = c(-LOGFC_CUTOFF, LOGFC_CUTOFF),
               linetype = "dashed", color = "grey40") +
    geom_text_repel(data = labels, aes(label = Gene),
                    size = 2.7, max.overlaps = 30, segment.color = "grey60") +
    facet_wrap(~ Compartment, ncol = 3) +
    labs(title = paste0("Compartment-stratified volcano: ", contrast_label),
         x = "log2FC", y = "-log10(FDR)") +
    theme_bw(base_size = 11) +
    theme(legend.position = "top", strip.text = element_text(face = "bold"))

  print(p)
  ggsave(file.path(subcell_dir,
                   paste0("Volcano_compartments_", contrast_label, ".pdf")),
         p, width = 12, height = 10, device = cairo_pdf)
}

for (cl in unique(msstats_ready$Label))
  plot_compartment_volcano(cl)
```

## Secretome candidates — flagged separately

Significantly changed proteins annotated as Extracellular / Secreted are useful biomarker candidates. Exported as a single ranked table.

```{r secretome}
secreted_genes <- compartment_map %>%
  filter(Compartment == "Extracellular / Secreted") %>%
  pull(Gene_upper)

secretome_table <- msstats_ready %>%
  filter(Status != "Not Significant", Gene_upper %in% secreted_genes) %>%
  arrange(Label, adj.pvalue) %>%
  select(Label, Gene, Protein, Description, log2FC, adj.pvalue, Status)

cat("Secreted/extracellular hits:", nrow(secretome_table), "\n")
fwrite(secretome_table, file.path(subcell_dir, "Secretome_candidates.csv"))

head(secretome_table, 30)
```

## Per-compartment top hits — wide table per contrast

For each contrast and compartment, the 10 most significant up- and down-regulated proteins.

```{r top_per_compartment}
top_compart <- sig_compart %>%
  arrange(Label, Compartment, adj.pvalue) %>%
  group_by(Label, Compartment, Status) %>%
  slice_head(n = 10) %>%
  ungroup() %>%
  select(Label, Compartment, Status, Gene, log2FC, adj.pvalue)

fwrite(top_compart, file.path(subcell_dir, "Top10_per_compartment_per_contrast.csv"))
head(top_compart, 25)
```

## Session info

```{r}
sessionInfo()
```
